### 2015-06-29 markdown/manuscript.tex
final revisions

\section{Why}\hypertarget{why}{}\label{why}

More than a decade ago, Rick Riolo, Bill Worzel and I were working on a consulting project together that involved genetic programming. As we chatted one day, Rick was asked what he'd most like to see as part of the research program of GP ``moving forward''. Rick's answer informs this contribution, as well as much of my professional work with GP in the years since. As I recall, he said he'd like work to focus on the ``symptoms'' we often see in evolutionary search processes: premature convergence, slow and spotty improvement, the catastrophic lack of diversity, and that ineffable feeling all GP professionals experience when they see results and \emph{know} that somewhere, somebody has Chosen Unwisely.

The literature abounds with well-written papers describing tips for avoiding local minima, improving on common search operators, and describing ``horse races'' between Bad Old and Better New search methodologies applied to benchmark problems. But I take Rick's challenge not only to mean that it would be useful to have a catalog which lists situations where search operator $X$ acting under contingency $Y$ tends to produce outcome $Z$, but also that \emph{the things we identify as symptoms themselves} are poorly understood.

This contribution aims squarely at an unfortunate sensibility I've detected in our field, which sees the many ``symptoms'' and ``pathologies'' as bugs to be eradicated \emph{before} we let ``real users'' run their own GP systems. Instead I'll argue (belatedly following Rick's lead) that these ``surprises'' and ``disappointments'' are not only inevitable but should be the main focus of our theoretical and practical work. Indeed, much of the value GP can offer can be lost if we approach these phenomena as something to be ``cleaned up'' rather than being \emph{accommodated}.

\subsection{Challenges}\hypertarget{challenges}{}\label{challenges}

Somewhere during that same project years ago, I remember a Project Manager recounting the story that when domain expert customers were shown a collection of Pareto-optimal solutions to the problem being explored, they were upset and confused: ``We don't want a choice, we want \emph{the best}.'' Of course there could be no \emph{best}, and they supposedly knew that; this response came after \emph{months} of analysis, interviews, technical planning and a collective agreement that the problem was fundamentally multi-objective. Even so, when the decision-makers were faced with the \emph{unquestionably} successful results they had collectively specified and specifically asked for\ldots{} those results weren't right enough.

In hindsight, I've come to think this shocked response was inevitable. It wasn't an unwitting surprise from some silly lay ``customers'' in an application project. Indeed, it's analogous to the way many in our field treat \emph{interesting} GP projects, whether they're ``theoretical'' or ``practical'', ``small'' or ``large''. \emph{Any interesting project will resist our initial plans and expectations.} When we're faced with even a little resistance, our first reaction is to imagine something has ``gone wrong in the setup'', shut the thing off and start again with ``better'' parameters.

It's unusual for anybody to consider interrogating the system \emph{in itself} when it resists, and adapt or accommodate that resistance. Within our field, but also in a surprisingly broad layer of technical society.

The field of GP has grown quite a bit in the 15 years since my two anecdotes. We who follow it closely have watched as frameworks, techniques and methodologies have diversified. There are the extraordinary ones we describe to one another in our little workshops, but also an increasing number of newsworthy and admirable public successes.

But this acceleration of innovation brings with it a dilution of the theoretical warrants we use to justify results. Most practical successes today highlight unique domain-specific quirks. New language and framework implementations makes numerous contingent (and often arbitrary) design and architectural choices. Even the fundamental ontological concepts of ``individual'', ``fitness'', ``behavior'', ``generation'' and ``population'' start to get troubling when try to fit many modern algorithms into their categories.

This growing divergence between current theory and practice has consequences both outside and within our field. The continued lack of interest in GP among statisticians, planners, designers and fans of traditional mathematical programming models is not merely a matter of disciplinary border wars. As we increasingly lose the ability to say \emph{why} something is happening in a given GP project, we are faced with an audience who stopped paying attention to progress in GP last century. As a result, students and researchers aiming to convince peers outside the field are forced to undertake mismatched experiments and apply unconvincing statistical tests. In response to these constant distractions, much time and attention goes towards unhelpful work guaranteeing that which cannot be guaranteed: running ``replicates'' of a process that is \emph{designed} to provide novel answers, measuring ``reliability'' of a process that \emph{intentionally} skirts dynamical chaos, providing ``summary charts'' of a process that strives to be as complex as the evolution of living beings, and (possibly worst) setting strict deadlines for an open-ended process to ``finish'' or ``succeed''.

I hasten to say this is not a fault in our field, but rather a broader philosophical and cultural problem. But it is nonetheless a misconception that undermines our understanding of what GP is \emph{doing}, of the way it unfolds in theory and in practice, and even what it's \emph{for}.

\subsection{A top-down approach}\hypertarget{a-top-down-approach}{}\label{a-top-down-approach}

I will make my case here in the form of an exercise for advanced GP users, or \emph{kata}. The habit of pursuing \emph{kata}, ``code retreats'', ``hackathons'' and other skill-honing practices is popular among software developers, and especially among the more advanced. Indeed, the title of my exercise (``GP as if you meant it'') is taken from a particularly influential exercise designed by Keith Braithwaite, ``TDD as if you meant it''.

Braithwaite's exercise feels subjectively \emph{harder} for more advanced programmers honing their development form; he suggests that novice programmers haven't learned ingrained but questionable habits, and haven't identified ``shortcuts'' that ``simplify'' the practices. In the same way, this exercise will feel \emph{most} artificial and restrictive to those of us with the most experience with GP. But like the martial arts exercises from which software \emph{kata} first were inspired, it isn't intended to be simple or even pleasant for the participants.

I intend it to expose habits I'll tactlessly call ``pseudo-GP'' among those of us who have come to think it's cheap and painless to \emph{just shut it off and start over} when things start to get strange in the course of a GP project. As an important side-effect, it helps surface that philosophical problem I alluded to above. But I have found the most useful result is how strongly and immediately it suggests tools with which one can address Rick's years-ago wish. It forces the participant to formally identify ``pathology'' and ``symptom'' before allowing them to attempt a ``fix''\ldots{} and even that must done with an intentionally limited set of tools.

Note, what I'll describe is a ``thought-experiment'', nor is it a serious suggestion for a new way of working on ``real problems'', nor as ``training'' for newcomers to the field. Rather it is designed as a rigorous and formal exercise to be undertaken by those of us already working closely with GP systems. Its intent is to surface the three shortcomings I've identified above:

\begin{enumerate}
\item what GP \emph{does}
\item how GP \emph{unfolds}
\item what GP is \emph{for}
\end{enumerate}

@@

As far as I can tell, Keith Braithwaite first described his \href{http://cumulative-hypotheses.org/2011/08/30/tdd-as-if-you-meant-it/}{training exercise} for software developers in 2009. His target was a sense that software developers who thought they were using test-driven development practices were in fact doing something more like ``Pseudo-TDD'', a sort of slapdash and habitual approximation lacking many of the benefits of mindful practice.

@@

Throughout the exercise, a facilitator patrols teams of participants and deletes \emph{any and all code not called for by a pre-existing failing test}. Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens, as one might imagine. But as Gojko Adzik emphasizes in descriptions of workshops he's facilitated, the resulting designs even for well-known algorithms in this artificially amplified setting become much more ``open-ended'' than would be expected if the code were written under the offhand attention of an experienced programmer without painful constraints.

Adzik mentions something in his account of a Tic-tac-toe exercise that's especially interesting:

@@

\section{GP as if we meant it}\hypertarget{gp-as-if-we-meant-it}{}\label{gp-as-if-we-meant-it}

In the same way that Braithwaite's coding exercise uses an onerous extra constraint\footnote{In this the sensibility reminds me of the constraint-driven art collective \href{http://en.wikipedia.org/wiki/Oulipo}{Oulipo}, who are perhaps most famous for the \emph{lipogram}, a literary work which cannot use a particular letter of the alphabet.} to drive participants towards more mindful and insightful decision-making, in this exercise I will demand a \emph{warrant} for each implementation decision that moves a running GP setup away from random guessing. Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function and \emph{post hoc} analysis is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed} in the course of the run.\footnote{Braithwaite's participants often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong''. I imagine many GP users will say they \emph{know} and \emph{use} all the innumerable design and setup options of GP, but treat them as adjustments to be invoked only when ``something goes wrong''. I offer no particular justification for either anecdote here, but the curious reader is encouraged to poll a sample of participants at any conference (agile or GP).}

Not only will traditional search operators like crossover, mutation and [negative] selection not come ``for free'' in this variant, but in every case we must develop a cogent argument in favor of starting them \emph{as part of an ongoing search process}. Similarly, the initial selection criteria will be limited to a minimal single element of the training data, and expansion of the training set in use will have to be made in light of measured progress, not the assumption that ``more is better''.

The result is a painfully incremental process of refinement of an unstoppable search which has \emph{intentionally} ``started wrong'', which must be carried out not by stopping the whole thing and launching it again with parameter ``tweaks'' but rather by \emph{doing surgery on the living patient} to correct any perceived ``pathologies''. Along the way, a fraction of the mysteries of ``pathology'' become much clearer and well-defined, even when they cannot be solved easily.

Again, let me emphasize that this is not intended as an ``algorithm'' in its own right, but rather as a suite of constraints imposed on what we do without paying attention already.

\subsection{The game tableau}\hypertarget{the-game-tableau}{}\label{the-game-tableau}

The exercise is structured as a sort of game with two players, one of which is a running GP System, and one which is a human User (or team of Users collaborating on making a single move at a time). The two players make alternating turns.

The game ``setup'' can be imagined as a \emph{tableau} with two components: A single list of search {\tt operators}, and a large (and growing) two-dimensional spreadsheet-like table with {\tt answers} as its row labels and {\tt rubrics} as its column labels.

During the User's turn, she can examine the state of the tableau, including all of its history, and can do any amount of data analysis or statistical consideration she wants. She is however permitted no more than \emph{two} moves: in each turn she can code and launch one new {\tt operator}, and can code and launch one new {\tt rubric}.

Finally, the User is obliged to provide a convincing \emph{warrant} for every move she make, to be reviewed by an external Facilitator. In the event this warrant is seen by the Facilitator to justify the particular move, he has the right to delete all {\tt operators} and {\tt rubrics} added by the User from that point forward \emph{without resetting the System's moves}.

Facilitator buy-in for every move is therefore \emph{strongly} encouraged.

During the System's turn, it will execute a finite number of steps in which it creates new individuals using the specified search {\tt operators} and {\tt rubrics}. At the end of the System's turn, control reverts to the User, and vice versa.

@@

An {\tt answer} might traditionally be called an ``individual'' in the GP literature. It is modeled here as key--value hash (or a simple prototypal object). Every new {\tt answer} is born with only a unique {\tt id} and its {\tt script} field set. As they ``mature'' during the System's turns, various other attributes (``scores'') will be set by {\tt rubrics}.

Answers never ``die'', and cannot be removed from the tableau by either player as search progresses. The User can only reduce or increase the chance that certain types of {\tt answer} contribute to later generations by constructing and launching {\tt rubrics} to adjust selective forces.

@@

An {\tt operator} is a coded function which takes as its argument an unordered  collection of {\tt answers}, and which produces a new collection of {\tt answers} as output.

The initial tableau includes only a single {\tt operator}, which implements a pre-coded ``random guessing'' algorithm. On the User's move, she may build and launch other more complex (and familiar) {\tt operators} like ``crossover'' or ``mutation'' functions. A wide variety of {\tt operator} structures are possible in the domain-specific language provided.

The specific set of {\tt answers} to which an {\tt operator} is applied cannot be chosen directly by the User. All ``parents'' defined for each {\tt operator} are chosen by \emph{lexicase selection} automatically during the System's turns, using the complete suite of {\tt rubrics} in play when invoked. This selection process samples each {\tt rubric} in the tableau with equal probability.

It is feasible for the User to design a new ``selection'' {\tt operator}, but it must rely on the immediate state of the {\tt answers} it is given; in other words, no {\tt operator} can explicitly invoke a particular {\tt rubric} (or vice versa).

Note again that no mechanism exists which removes {\tt answers} from the tableau.

@@

A {\tt rubric} is a function which returns a scalar value (not necessarily a \emph{number}) for a given {\tt answer}, conditioned as needed on the instantaneous tableau state. When a {\tt rubric} is applied to an {\tt answer}, it sets a new value (or ``score'') in that {\tt answer}.

No ``score'' is ever updated in an {\tt anwer} once set by a {\tt rubric}. However, if multiple {\tt answers} exist which have the same {\tt script}, each may be scored at different times or in different contexts, and the resulting values may differ.

An aggregate or \emph{higher-order} {\tt rubric} can be created, but as a consequence it ``entails'' all of its component {\tt rubrics}. So if the User constructs a higher-order {\tt rubric} like ``max error over 30 training cases'', when she launches that {\tt rubric} 31 in total will be launched, with the first 30 being in the form ``absolute error for training case $i$''.

If a User launches such a higher-order {\tt rubric} on her turn, all its entailed {\tt rubrics} will also be added before the System takes its turn.

When one of these ``input'' {\tt rubrics} is applied during the System's turns, it does \emph{not} automatically cause the higher-order aggregate to be applied. However, when the higher-order {\tt rubric} is applied, it will force all of its ``input'' {\tt rubrics} to be calculated and applied first.

In addition to ``entailed'' {\tt rubrics} for aggregate scoring, the User can use any of the other extant {\tt rubrics} present in the tableau in her definition, or the state of the interpreter before and after the {\tt script} is run. It is therefore entirely possible to specify functions which score:

\begin{itemize}
\item number of characters in the {\tt answer}'s script (since the {\tt script} is an attribute of every {\tt answer} with its own column in the tableau, and therefore its value is exactly equivalent to any other {\tt rubric}-created score)
\item maximum error measured in any of 35 other {\tt rubrics}
\item number of {\tt div0} errors produced when running with a particular set of inputs
\item number of stochastic instructions appearing in the {\tt answer}'s script, compared to \emph{all other {\tt answers}}
\end{itemize}

Various problem-specific aspects are glossed here. Suffice to say, the construction of reasonable {\tt rubrics} in response the System's moves is the core of the User player's game strategy.

\subsubsection{User turn}\hypertarget{user-turn}{}\label{user-turn}

In her turn, the User can do either or both of the following:

\begin{enumerate}
\item create and launch one {\tt operator}
\item create and launch one {\tt rubric} (which can be ``higher-order'' and entail others)
\end{enumerate}

To support these decisions, the User can examine the tableau state and history in detail. Before the decisions are implemented, though, a \emph{warrant} must be written for each one. Whether the Facilitator is asked to review this is entirely up to the User player, subject to the warnings mentioned above.

\subsubsection{System turn}\hypertarget{system-turn}{}\label{system-turn}

During its turn, the System player adds a specified minimum number of new {\tt answers} to the tableau, one at a time, using a simple form of lexicase selection. Until it reaches its halting state it iterates this cycle:

\begin{enumerate}
\item select one {\tt operator} from those in the tableau, with equal probability
\item apply \emph{lexicase selection} to the tableau to select the required number of input {\tt answers}, filling in missing {\tt rubric} scores as needed
\item apply the {\tt operator} to the inputs to produce one or more new {\tt answers}, and append those new {\tt answers} immediately to the tableau
\item {\tt HALT} if the number of new {\tt answers} meets or exceeds the limit; otherwise, go to step (1)
\end{enumerate}

\subsubsection{Choosing a language}\hypertarget{choosing-a-language}{}\label{choosing-a-language}

No particular constraints apply to the language or representation chosen for the exercise, except that it should be sophisticated enough to support \emph{redundant capacity}. That is, for any given algorithmic goal (or stage), there should be multiple paths to success.  So for example if the problem could reasonably be expected to involve ordered lists, then it would be wise if the language had at least two different ways to ``deal with lists'': iterators, recursion, a comprehensive set of second-order functional operators, an explicit {\tt List} type with associated methods, that sort of thing.

If the language is sophisticated enough to have ``libraries'' of instructions and types intended for specialized domains, I'd strongly encourage that \emph{all of these should be used}. That is, the facilitator should err on the side of ``winnowing complexity'', rather than forcing the User and System to \emph{invent basic data structures} or the idea of floating-point numbers at the same time they're trying to solve the ``real'' problem under consideration.

>>>>

\section{Why: An Excuse}\hypertarget{why-an-excuse}{}\label{why-an-excuse}

More than a decade ago, Rick Riolo, Bill Worzel and I worked together on a genetic programming consulting project. As we chatted one day, one of them---I don't recall which, and accounts vary---was asked what he'd most like to see as GP ``moved forward'', and said he'd want the field to think more about the ``symptoms'' we so often see when we apply evolutionary search processes in complex settings. That is: premature convergence, slow and spotty improvement, a catastrophic lack of diversity, and more generally that ineffable feeling all GP professionals experience when we look at results and \emph{know we have chosen unwisely}.

Now the reader will point out that the literature in our field overflows with well-written papers describing tips for avoiding local minima, improving on common search operators, and running ``horse races'' between Bad Old and Better New search methodologies applied to benchmark problems. But while as a rule these are presented as a sort of \emph{general principle}, most in practice are \emph{case studies} in which it is shown, for example, that search operator $X$ acts under contingency $Y$ and sometimes produces outcome $Z$.

I am not left with a sense that this catalog addresses my colleague's stated wish. While it is surely necessary to compile such a list of individual observations under particular suites of experimental treatments and benchmarks, it is insufficient until we can achieve the skill of recognizing \emph{when} something noteworthy is happening. In particular, I want to focus on those frequent but contingent situations in which we are willing to say our GP system \emph{resists our expectations}.

A good deal of this chapter will be spent explaining just what I mean by this  particular usage of ``resistance'', but for the moment the idea will be clear enough from this simple mental exercise: What are things GP has ``done''---in the context of a particular project---that have left you feeling unsatisfied, confused, or frustrated? Not because you've made a simple mistake setting a parameter, or introduced a bug in a codebase, but because (at that moment) you have no idea why the GP process is doing \emph{that thing} under those particular circumstances? Maybe it's not converging when you know it should; maybe it's exploring unexpectedly complex algorithms rather than obvious simpler ones, despite your reasonable parameter settings; maybe it's failing to solve simple problems but easily solving hard ones. In any case, the system ``resists'' your careful \emph{and knowledgeable} expectations by inexplicably refusing to follow your plan.

My general point, and the particular point of this chapter and the exercise it describes, is that we should then ask: \emph{When we feel GP is resisting us, what should we do---and why?} Further, after more than twenty years of work in the field, I think we have learned enough for this to be reasonable and productive research program.

I'm sure every reader has at least one story they can tell, in which GP has ``resisted'' in this sense. I want to draw our attention to these incidents from a conviction that it is exactly the steps we take to \emph{accommodate} GP's resistance which provoke our sudden insights into the structure of our problem, drive us to build a ``workaround'' that becomes a novel selection algorithm or architecture, and lead us to successfully re-frame our project to use a completely new approach. Almost any chapter in this series of GPTP Workshop Proceedings will contain a story just like this: We create an age-layered population because working without age-layering didn't perform as expected \citep{Hornby:2009:GPTP}; we invent a new selection mechanism because traditional algorithms worked as described, but failed to capture crucial details of our problems \citep{Spector:2012:GECCOcompA}; we create an entirely new representation and suite of search operators because our goals aren't met by snipping up and recombining in the traditional way \citep{ryan:2003:GPTP}; and so on.

The point in each case is: one gains little \emph{insight} into a problem when GP quickly pops out the ``right answer'' without a fight. Too often I see papers treating this ubiquitous resistance as something to be eradicated \emph{before} ``real users'' are allowed run their own GP searches. Instead I'll argue here that ``surprises'' and ``disappointments'' are not only inevitable but \emph{are the main source of value in many GP projects}, and as such should be the main focus of our theoretical and practical work.

Every interesting project will resist our plans and expectations. But in our field, whenever we're faced with even a bit of this sort of behavior, our first reaction is generally to act as though something has ``gone wrong in the setup'', then shut the misbehaving run down so we can start again with ``better parameters'' next time. I find it unusual for anybody to interrogate the system \emph{in itself} when it resists, or to make an attempt to adapt or accommodate perceived resistance. I hasten to say this is not a fault with our field, but rather a symptom of broader philosophical and cultural problems in our approach to programming and computational research projects, and our understanding of computing more generally.

The exercise I describe in this chapter is intended to bring our attention back on GP as a dynamical process \emph{in itself}, as opposed to a tool to be adjusted ``in between'' applications. It may be the case that other ``traditional'' machine learning methodologies are built on better-defined information-theoretic foundations, and come with suites of strong statistical tests for overfitting and robustness; as a result it's perfectly reasonable to treat them as \emph{tools}, and as \emph{malfunctioning} or just being \emph{wrong for a problem} when they act in unexpected ways. But GP is somehow a different sort of animal: when it ``resists'', we are left with an essentially unlimited choice of how we should \emph{accommodate} that resistance.

\subsection{On mindful exercises}\hypertarget{on-mindful-exercises}{}\label{on-mindful-exercises}

The habit of pursuing \emph{kata}, ``code retreats'', ``hackathons'' and other skill-honing practices is popular among software developers, and especially among the more advanced. It apparently arose independently among a few groups in the 1980s, but Dave Thomas seems to have first used the Japanese term \citep{Thomas2013}.

Indeed, the title of my exercise (``GP As If You Meant It'') is directly inspired by from an exercise designed by Keith Braithwaite, ``TDD as if you meant it'' \citep{Braithwaite2011}. One of the interesting aspects of Braithwaite's exercise is that it feels subjectively ``harder'' when attempted by advanced programmers honing their development form; he suggests that novice programmers haven't learned ingrained but questionable habits, and haven't identified ``shortcuts'' that ``simplify'' the practices.

In the same way, the exercise I describe will feel \emph{most} artificial and restrictive to those of us with the most experience with GP. But like the martial arts exercises by which software \emph{kata} first were inspired, it isn't intended to be simple or even pleasant for the participants. Just as Braithwaite's exercise targeted what he calls ``Pseudo-TDD'', I intend mine to surface habits we can think of as ``pseudo-GP'': the sense that it's cheap and painless to \emph{just shut it off and start over} when we start sensing a problem is arising.

\subsection{Caveats}\hypertarget{caveats}{}\label{caveats}

What I describe here should be considered a ``thought-experiment'' backed up by my own experiences; there are only a few sketchy manual implementations to date. Some readers have imagined that the ``automated'' System player must somehow be a self-contained and tested process running on the cloud (and that is an eventual goal), but in all my early tests that role has been played by a handful of simple Ruby scripts, edited between ``turns'' by the Facilitator to reflect changes made by the User player. In other words, \emph{there need not be an actual autonomous ``computer player'' in this game}, unless you feel compelled to write one up beforehand, and the game can as easily be played by a Facilitator who \emph{codes} the moves of the System and \emph{plays that role}, plus the User as described.

Further, there is no reason the User player need be a single human being. Indeed, it's common practice in many software development \emph{katas} to work in teams or even as an aggregated crowd of everybody in the room. Many hands make light work, especially on a tight schedule. In other words, these roles are intended to be ``notional'' rather than definitional.

I should also point out that this game is not a serious suggestion for a new way of working on ``real problems'', nor as any sort of ``training'' for newcomers to the field of GP. Rather it is designed as a rigorous and formal exercise in mindful practice, to be undertaken by people already working closely enough with GP systems to recognize the problems when they arise.

@@

Keith Braithwaite seems to have first described this training exercise for software developers in 2009 \citep{Braithwaite2011}. His target was a sense that software developers who thought they were using test-driven development practices \citep{Beck2002} were in fact doing something more like ``Pseudo-TDD'', a sort of slapdash and habitual approximation lacking many of the benefits of mindful practice.

@@

Throughout the exercise, a facilitator patrols teams of participants and deletes \emph{any and all code not called for by a pre-existing failing test}. Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens, as one might imagine. But as Gojko Adzic \citep{Adzic2009} has said in descriptions of workshops he's facilitated, the resulting designs even for well-known algorithms in this artificially amplified setting become much more ``open-ended'' than would be expected if the code were written under the offhand attention of an experienced programmer without painful constraints.

Adzic passes along some observations in his account  of a Tic-Tac-Toe exercise \citep{Adzic2009} that are especially interesting for me:

@@

\section{GP As If You Meant It}\hypertarget{gp-as-if-you-meant-it}{}\label{gp-as-if-you-meant-it}

In the same way that Braithwaite's coding exercise uses an onerous extra constraint\footnote{In this the sensibility reminds me of the constraint-driven art collective Oulipo \citep{Becker:2012}, who are perhaps most famous for the \emph{lipogram}, a literary work which cannot use a particular letter of the alphabet.} to drive participants towards more mindful and insightful decision-making, in this exercise I will demand a \emph{warrant} for each implementation decision that moves a running GP setup away from random guessing. Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function and \emph{post hoc} analysis is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed} in the course of the run.\footnote{Braithwaite's participants \citep{Braithwaite2012} often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong''. I imagine many GP users will say they \emph{know} and \emph{use} all the innumerable design and setup options of GP, but treat them as adjustments to be invoked only when ``something goes wrong''. I offer no particular justification for either anecdote here, but the curious reader is encouraged to poll a sample of participants at any conference (agile or GP).}

Not only will traditional search operators like crossover, mutation and [negative] selection not come ``for free'' in this variant, but in every case we must develop a cogent argument in favor of starting them as part of an ongoing search process. Similarly, the initial selection criteria will be limited to a  single training case, and expansion of the active training set will have to be made \emph{in response to particular features} of observed progress, not merely on the basis of the assumption that ``more will be better''.

The result is a painfully incrementalized process, one that focuses on the refinement and eventual correction of an unstoppable search which was \emph{intentionally} ``started wrong'', and which must be carried out by \emph{doing surgery on the living patient} to correct perceived ``pathologies'' and ``resistance''. Along the way, a fraction of the mysteries of ``pathology'' has the potential to be much clearer and better-defined.

\section{Overview}\hypertarget{overview}{}\label{overview}

The exercise is structured as though it were a game for two players, plus a Facilitator who establishes the ground rules, provides any needed technical infrastructure, and acts as referee. One player is (or represents) a running GP System, and one player is (or represents) a human User trying to mindfully drive the System's performance in a desirable direction over a series of turns.

While I speak below of the System player ``being'' a self-contained software process, it is of course a loose role that might more easily be played by another human, potentially the Facilitator herself, writing and running a simple series of scripts on a laptop. Similarly, while I may say that the User player ``is'' a single human being, it could as easily be a room-full of students or workshop participants, or a mailing list voting over many weeks on strategies for each turn. Indeed, in working out the exercise as it's described here, I've ``played'' all the roles myself, simultaneously, and still found interesting and unexpected insight.

In preparing the \emph{kata}, the Facilitator selects a \emph{target problem}, which should be a supervised learning task for which plenty of data is available. The target should not be ``toy'' in the sense of having a simple, well-known answer; rather it should be challenging and open-ended enough to warrant a publication if solved (a problem that has \emph{recently} been solved might do in a pinch, though there is no shortage of open ones). Any good book of mathematical recreations (for example, \citep{Winkler:2003} or nearly any book by Martin Gardner, Ian Stewart or Ivan Moscovich) will provide numerous abstract problems that have never been attempted with GP.

The Facilitator will also need to choose a representation language, set up an initial Tableau, and provide enough software infrastructure so that the System's turns can be made easily. In each turn of the game, both the System and the User players take actions to modify and extend the initial Tableau: the User's options include adding \emph{rubrics} and \emph{operators} (described below), and in its turns the System acts ``mindlessly'' by \emph{invoking} the specified rubrics and operators to add a fixed number of new individuals to the growing population.

The User's goal is to drive the System towards producing \emph{sufficiently good solutions} to the target problem, but their decisions are constrained by the obligation to provide a \emph{warrant} for every change made which is convincing to the Facilitator. The particular definition of a ``sufficiently good solution'' is left for the players and Facilitator to decide in context: it may be impossible to completely ``solve'' some target problems, given the tools at hand.

It should always be kept in mind that \emph{the point of the exercise is to evoke interesting and useful warrants}, not merely to drive the System in a desired direction. In each turn the User is obliged to produce a convincing \emph{warrant} for every move she makes, which must be reviewed and approved by the Facilitator before play proceeds. These warrants need not be \emph{factually correct}, but they must be \emph{convincing} in the context of the game state at the time they're put forth.

\subsection{The Tableau}\hypertarget{the-tableau}{}\label{the-tableau}

The game ``board'' is a \emph{Tableau} with two components: A list of search {\tt operators}, and a two-dimensional spreadsheet-like table which uses {\tt answers} as its row labels and {\tt rubrics} as its column labels. Initially the Tableau is empty, except for a single entry on the list of {\tt operators} labeled ``random guess''.

The ``random guess'' operator is set up by the Facilitator before play begins; it produces a single random {\tt answer} for the target problem, with no arguments. Think of it as equivalent to the ``initialization'' function used to produce a single ``random'' individual in a traditional GP setting.

During the User's turn, she can examine the state of the Tableau, including any or all of its history, and the algorithms in play, and apply any amount of data analysis or statistical work she wants. On her turn she is permitted no more than \emph{two} moves: She can (optionally) code and append one new {\tt operator} to the list of {\tt operators}, and she can (optionally) code and append one new {\tt rubric} column to that table.

During the System's turn, it will produce a fixed number of new {\tt answers}. It creates each new {\tt answer} by first picking an entry from the list of {\tt operators} with uniform probability. When an {\tt operator} is chosen, it executes the indicated algorithm, selecting parents from the {\tt answers} table as needed. For each required selection of a parent, the scores recorded in the table of {\tt answers} vs {\tt rubrics} are used (see below). Once parents are selected, the {\tt operator} is applied and a new {\tt answer} is immediately appended to that part of the Tableau.

@@

What I'm calling an {\tt answer} here would probably be called an ``individual'' in the GP literature. In this case, it is a particular \emph{script} or program in the representation language the Facilitator has chosen for the target problem.

Answers never ``die'', and cannot be removed from the Tableau by either player.

@@

Each {\tt operator} is a function which takes as its argument an unordered  collection of zero or more {\tt answers}, and which produces a new collection of one or more {\tt answers} as output.

The initial Tableau includes only a single {\tt operator}, which implements a pre-coded ``random guessing'' algorithm. On the User's move, she may build and launch other more complex (and familiar) {\tt operators} like crossover or mutation. The only permitted argument is a set of zero or more {\tt answers}; no numerical or other parameters are allowed. A wide variety of {\tt operator} algorithms are still possible, and the details of the code in which they are implemented is left to the Facilitator to specify as part of the game setup.

Except for some unusual edge cases, the \emph{specific} set of {\tt answers} to which an {\tt operator} will be applied cannot be chosen directly by the User. All ``parents'' defined for each {\tt operator} are chosen independently (and incrementally) by \emph{lexicase selection}, using the suite of {\tt rubrics} in play when the System takes its turn. This lexicase selection process samples every {\tt rubric} in the Tableau with equal probability. It may be possible for the User to design a new ``selection'' {\tt operator} which takes as its argument ``all the answers'' and somehow culls that collection down to a subset---but to do so, it must rely on the immediate state of the {\tt answers} it is given; {\tt rubric} scores are ``stored'' only in the Tableau {\tt answer} table itself, not as gettable attributes of the {\tt answers}, making it very difficult for any new {\tt operator} algorithm to use {\tt rubric} scores in its implementation.

Note again: no mechanism exists which \emph{removes} {\tt answers} from the Tableau.

@@

A {\tt rubric} is a function which returns a scalar value (not necessarily a \emph{number}) for any given {\tt answer}, conditioned (as needed) on the instantaneous Tableau state. When a {\tt rubric} is applied to an {\tt answer}, it sets a new value (or ``score'') for that {\tt answer} in the appropriate cell of the ``spreadsheet'' portion of the Tableau.

No score is ever changed for an {\tt answer}, once it has been set by a {\tt rubric}. However, if multiple {\tt answers} exist which have the same {\tt script}, each may be scored at different times or in different contexts, and the resulting values may differ.

An aggregate or \emph{higher-order} {\tt rubric} can be created, but its existence ``entails'' all of the component {\tt rubrics} from which its score is derived. So, for example, if the User constructs a {\tt rubric} like ``maximum absolute error observed over any of these 30 training cases'', on the turn when she adds that {\tt rubric} up to 31 new columns will be added to the spreadsheet portion of the Tableau: one for each of the 30 requisite ``sub-{\tt rubrics}'', which produce as their scores the absolute error for a single training case, and also one \emph{aggregate} {\tt rubric} which calculates and reports ``maximum of those other 30 scores''. Whenever the User submits such a higher-order {\tt rubric} on her turn, all of the entailed {\tt rubrics} are added automatically if they do not already exist. Note however that there must still be a \emph{warrant} for the {\tt rubric}, which convinces the Facilitator of its potential usefulness.

The User can use any information present in the Tableau in building new {\tt rubric} functions (such as row numbers, string values of {\tt answers}, or scores), and can also examine the detailed state of the interpreter before and after the {\tt script} is run. It should be clear therefore that the User can specify {\tt rubric} functions which score aspects of the problem such as:

\begin{itemize}
\item the absolute error for a single training case
\item the number of tokens in the {\tt answer}'s script
\item maximum error measured in any of 35 other {\tt rubrics}
\item number of {\tt div0} errors produced when running a script with a particular set of inputs
\item number of stochastic instructions appearing in the {\tt answer}'s script
\item rewrite difference between the scored script the most common {\tt answer} in the population
\item (and so on)
\end{itemize}

Various problem-specific aspects are glossed here, and it is left to the Facilitator to be reasonable in the context of the target problem and the representation she has chosen. Suffice to say, the construction of useful {\tt rubrics} in response the System's moves is the core of the User player's game strategy.

\subsection{Lexicase Selection}\hypertarget{lexicase-selection}{}\label{lexicase-selection}

All {\tt operator} inputs are chosen by lexicase selection. Slightly simplifying Spector's original description \citep{Spector:2012:GECCOcompA}, the following algorithm can be used:

\begin{itemize}
\item (beginning with a ``set under consideration'' which includes all {\tt answers}\ldots{})
\item for each {\tt rubric} in a random permutation of all the {\tt rubrics} in the Tableau, discard all {\tt answers} from the set under consideration whose score on this {\tt rubric} is sub-optimal (relative to the current set under consideration)
\item if multiple {\tt answers} remain after all {\tt rubrics} have been applied, return one {\tt answer} picked randomly, with uniform probability, from the remaining set under consideration
\end{itemize}

At the beginning of the game when no {\tt rubrics} have been added and no {\tt operators} exist which require ``parents'', no selection occurs or needs to occur. But note that the algorithm as described would \emph{still} provide a parent, even when no {\tt rubrics} have been specified: an empty set of {\tt rubrics} would be immediately exhausted without eliminating any {\tt answers}, and then a single {\tt answer} would be selected with uniform probability from that complete set.

Lexicase selection has several characteristics which argue for its use here; in particular it is interesting because it seems (anecdotally, but see also Helmuth in this volume) to be a relatively ``slow'' selection method, permitting a diversity of {\tt answers} to coexist in a population at the same time. However I admit readily that I've specified it as a core part of this \emph{kata} because it is \emph{unfamiliar to most prospective players}, and therefore more likely to produce unexpected behavior of a useful sort.

\subsection{The User's turn}\hypertarget{the-users-turn}{}\label{the-users-turn}

In her turn, the User can do either (or both) of the following:

\begin{enumerate}
\item add one {\tt operator} to the list in the Tableau
\item add one {\tt rubric} to the Tableau, which can be ``higher-order'' and therefore entail others
\end{enumerate}

To support these decisions, the User can examine the Tableau state and history in detail. Before any code is implemented, though, a \emph{warrant} must be written for each one.
